---
title: "Oral_Microbiome_LungCancer_Study"
author: "Vanessa DeClercq"
date: "2025-07-16"
output: html_document
---

## Metadata Analysis
-normality test 
-Descriptive analysis (counts, mean, median, SD, IQR) of covariates - sex, AGE, BMI, ETC.

Examples:
mean(metadata$age_at_surgery)
sd(metadata$age_at_surgery)
median(metadata$age_at_surgery)
IQR(metadata$age_at_surgery)
range(metadata$age_at_surgery)

```{r}
#check for normality
shapiro.test(metadata$age_at_surgery)

#counts of each gp
table(metadata$stage_cat)

#summary table of participants age by sex
library(PMCMRplus)
aggregate(age_at_surgery ~ Sex, data = metadata, summary)
kruskal.test(metadata$age_at_surgery, 
             metadata$Sex)
kwAllPairsDunnTest(metadata$age_at_surgery, 
                   as.factor(metadata$Sex), "BH")
t.test(metadata$age_at_surgery~metadata$Sex)

mean(metadata$age_at_surgery)
range(metadata$age_at_surgery)


#TABLES FOR COUNTS, PERCENTAGES - CATEGORICAL VARIABLES
#install.packages("gmodels")
library(gmodels)
library(reporttools)

Smoking_Status <- CrossTable(metadata$Smoking_Status, 
                        metadata$Sex, 
                        prop.r=TRUE, chisq = TRUE)
pairwise.fisher.test(metadata$SampleType, metadata$stage_group,
                     p.adjust.method = "BH")


```

##Diversity Analysis

#Alpha Diversity
Use QIIME2 generated alpha diversity metric. 
First, test for normality. Then perform parametric or non-parametric test for differences in alpha diversity among sample types. 
Four different alpha diversity metrics will be used; richness, evenness, Faiths PD, Shannon diversity.
```{r}
#Reading Alpha Diversity file
alpha <- read_csv("alpha_diversity.csv")
#match alpha diversity file to metadata file
intersect(rownames(metadata),rownames(alpha))
samples_to_keep <- intersect(rownames(metadata), rownames(alpha))
filtered_alpha <- data.frame(alpha[samples_to_keep,])
rownames(filtered_alpha) <- filtered_alpha$SampleID

#test for normality.
for (i in 2:5){print(shapiro.test(filtered_alpha[,i]))}
#results show none are normal so I will have to use non-parametric tests moving forward.

#use Kruska wallis test to test for significant differences bewteen sample types.
for (i in 2:5){print(kruskal.test(filtered_alpha[,i], filtered_metadata$stage_cat))}

#post-hoc test for between group differences.
library(PMCMRplus)
for (i in 2:5){print(kwAllPairsDunnTest(filtered_alpha[,i], as.factor(filtered_metadata$stage_cat), "BH"))}

###Additional alpha diversity analysis using Ordinal Logistic Regression
library(MASS)
#unadjusted
m1 <- polr(stage_cat ~ pielou_evenness,
  data = metadata, Hess = TRUE)
summary(m1)

#adjusted for smoking and antibiotic use
m2 <- polr(stage_cat ~ pielou_evenness + Smoking_Status + abx_6mon_prior,
  data = metadata, Hess = TRUE)
summary(m2)

#store table
(ctable <- coef(summary(m2)))

#calculate and sotre pvalues
p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2

## combined table
(ctable <- cbind(ctable, "p value" = p))

#confidince intervales
(ci <- confint(m2)) # default method gives profiled CIs

#OR and CI
exp(cbind(OR = coef(m2), ci))


##create box plot for alpha diversity measures
#boxplot with ggplot
 #merge the alpha diversity and metadata dataframes.
alpha_metadata <- merge(filtered_metadata, filtered_alpha, by = 'row.names', all=TRUE)
alpha_metadata <- data.frame(alpha_metadata, row.names = 1)

#ggplot
library(ggplot2)

ggplot(data=subset(alpha_metadata, !is.na(stage_cat)), aes(x=stage_cat, y=observed_features, fill = stage_cat)) +geom_point() + 
  ylim(c(0,370)) + 
  stat_boxplot(geom ='errorbar', width = 0.5) +
  geom_boxplot(alpha=1, color=1, outlier.shape = NA) +
  labs(x = "Cancer Stage", y = "Observed Features") +
  labs(fill = "Cancer Stage") +
  scale_fill_manual(values=c("#1B9E77", "#D95F02", "#7570B3")) +
  theme_bw() +
  ggtitle("Observed Features") +
  theme(plot.title = element_text(size = rel(1.5), face = "bold")) +
  geom_jitter(shape=16, position=position_jitter(0.2), size = 2) +
  theme(axis.title.x = element_text(size =16, face = "bold"))+
  theme(axis.title.y = element_text(size = 16, face = "bold", angle = 90))+
  theme(axis.text = element_text(size =14)) +
  theme(legend.title = element_text(size = 14, face = "bold"))+
  theme(legend.text = element_text(size = 14))

ggplot(data=metadata, 
       aes(x=OS_censored, y=observed_features, fill =OS_censored)) +geom_point() + 
  ylim(c(0,370)) + 
  stat_boxplot(geom ='errorbar', width = 0.5) +
  geom_boxplot(alpha=1, color=1, outlier.shape = NA) +
  labs(x = "Overall Survival", y = "Observed Features") +
  labs(fill = "EFS") +
  scale_fill_manual(values=c("#0072B2", "#D55E00")) +
  theme_bw() +
  ggtitle("Observed Features") +
  theme(plot.title = element_text(size = rel(1.5), face = "bold")) +
  geom_jitter(shape=16, position=position_jitter(0.2), size = 2) +
  theme(axis.title.x = element_text(size =16, face = "bold"))+
  theme(axis.title.y = element_text(size = 16, face = "bold", angle = 90))+
  theme(axis.text = element_text(size =14)) +
  theme(legend.title = element_text(size = 14, face = "bold"))+
  theme(legend.text = element_text(size = 14))


#correlation with alpha diversity
plot(filtered_metadata$age_at_surgery, filtered_alpha$observed_features, 
     xlab="Age (years)", ylab="Observed Features", 
     xlim = c(50, 90), ylim = c(0, 400), 
     pch=19, cex.axis = 1, cex.lab = 1.5, las=1)
abline(lm(filtered_alpha$observed_features ~ filtered_metadata$age_at_surgery),
       col = "blue", lwd=7)

test <- cor.test(filtered_alpha$observed_features, filtered_metadata$age_at_surgery, 
                 method=c("pearson"))    

test2 <- cor.test(filtered_alpha$observed_features, filtered_metadata$age_at_surgery,
                 method=c("spearman"))    


```

#Beta Diversity
#use of QIIME2 generated Bray Curtis dissimilarity matrix or Weighted UniFrac distance matrix.
# Bray Curtis
```{r}
#read in beta diversity files
bray_curtis <- read_csv("bray_curtis.csv")
bray_curtis <- as.data.frame(bray_curtis)
rownames(bray_curtis) <- bray_curtis[,1]

#match beta diversity data to metadata
intersect(rownames(filtered_metadata),rownames(bray_curtis))
samples_to_keep <- intersect(rownames(filtered_metadata),rownames(bray_curtis))
filtered_bray_curtis <- as.data.frame(bray_curtis[samples_to_keep,])
rownames(filtered_bray_curtis) <- filtered_bray_curtis[,1]

#adonis test - permutational ANOVA of dissimilarities
library(vegan)

set.seed(23)
adonis2(filtered_bray_curtis ~ filtered_metadata$stage_cat,
        permutations = 1000, by="margin")

#with covariates
set.seed(23)
adonis2(filtered_bray_curtis ~ filtered_metadata$OS_censored + filtered_metadata$OS_time,
        permutations = 1000, by="margin")


#plot beta diversity
library(devtools)
library(ggordiplots)
library(ggplot2)
bray_curtis_pcoa <- cmdscale(filtered_bray_curtis, k=2, eig = TRUE)
barplot(bray_curtis_pcoa$eig[1:10])
component1 <- bray_curtis_pcoa$eig[1]/sum(bray_curtis_pcoa$eig)
component2 <- bray_curtis_pcoa$eig[2]/sum(bray_curtis_pcoa$eig)
component1*100
component2*100

plot_data <- data.frame(pc1=bray_curtis_pcoa$points[ ,1],
                        pc2=bray_curtis_pcoa$points[ ,2],
                        Sample_type=filtered_metadata_bray$stage_cat)

bray_plot <- ggplot(data=subset(plot_data, !is.na(Sample_type)), aes(x=pc1, y=pc2, color=Sample_type)) +  geom_point(size=2) +
  stat_ellipse(linewidth=2) + scale_color_manual(values=c("#1B9E77", "#D95F02", "#7570B3"), 
                                            name="Stage", labels=c("I", "II", "III"))+
  ggtitle("Bray-Curtis") +
  theme_bw() + xlab("PC1(4.6%)") + ylab("PC2(3.3%)") + theme(legend.position = c(0.88,0.84)) +
  theme(text = element_text(size = 26)) + ylim(-0.5, 0.5)+ xlim(-0.5, 0.5)+
  theme(axis.text.x=element_text(size=rel(1.2)), axis.text.y=element_text(size=rel(1.2)))
bray_plot


```


##TAXA ANALYSIS


```{r}
#preprep data for analysis.
#read in taxa table & set row names)
taxa_table <- read_csv("taxa_table.csv")
taxa_table <- as.data.frame(taxa_table)
rownames(taxa_table) <- taxa_table[,1]

#create taxa table with only samples of interest, then flip the taxa table
#(filter taxa to match samples in the 'metadata' file).
intersect(rownames(taxa_table), rownames(metadata))
samples_to_keep <- intersect(rownames(taxa_table), rownames(metadata))
filtered_taxa <- taxa_table[samples_to_keep,]

#flip table so taxa are rows, samples columns
rare_filter_table <- t(filtered_taxa)

##MaAsLin2 DA analysis
library("devtools")
#install_github("biobakery/maaslin3")
library(maaslin3)
#install.packages("gtable")
library(gtable)

#unadjusted
set.seed(1)
fit_out <- maaslin3(input_data = filtered_taxa,
                    input_metadata = filtered_metadata,
                    output = 'Maalin3_output_stage',
                    formula = '~ stage_cat',
                    normalization = 'TSS',
                    transform = 'LOG',
                    reference = c('stage_cat,I'),
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 1)
#adjusted for covariates
set.seed(1)
fit_out <- maaslin3(input_data = filtered_taxa,
                    input_metadata = filtered_metadata,
                    output = 'Maaslin3_output_OS_censored_time',
                    formula = '~ OS_censored + OS_time',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 1)

#ANCOM2 DA analysis
library(compositions)
library(tidyr)
library(nlme)
library(dplyr)
library(ggplot2)
library(compositions)
library(exactRankTests)
deps = c("exactRankTests", "nlme", "dplyr", "ggplot2", "compositions")
for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(dep)
  }
  library(dep, character.only = TRUE)
}
source("~/Desktop/ANCOM2.R")


#preprocessing step - to deal with different types of zeros before performing differential abundance analysis 
preprocess <- feature_table_pre_process(feature_table = rare_filter_table, 
                      meta_data = filtered_metadata, sample_var = "Saliva_sampleID", 
                      out_cut = 0.05,zero_cut = 0.9, lib_cut = 1000,)

#run main ANCOM function with preprocessed data,adjusting p-values for multiple comparisons
# without covariates, use NULL
rez <- ANCOM(preprocess$feature_table, preprocess$meta_data, 
             preprocess$structure_zeros,"stage_cat", "BH", 0.1, NULL)
View(rez[[1]])
#with covariates
rez <- ANCOM(preprocess$feature_table, preprocess$meta_data, 
             preprocess$structure_zeros,"stage_cat", "BH", 0.1, "stage_cat + Sex")
View(rez[[1]])


#Corncob differential abundance (DA) analysis
library(corncob)
library(phyloseq)
library(magrittr)
#create the object containing the metadata and taxa to be tested.
otu_tab <- otu_table(rare_filter_table,taxa_are_rows = TRUE)
sam_data <- sample_data(filtered_metadata)
phylo <- phyloseq(otu_tab, sam_data)

#runs corncob DA analysis, returns plot and, DA results
results <- differentialTest(formula = ~ stage_cat,
                            phi.formula = ~ stage_cat,
                            formula_null = ~ 1,
                            phi.formula_null = ~ stage_cat,
                            data = phylo,
                            test = "Wald", boot = FALSE,
                            fdr_cutoff = 0.1)

results$p_fdr
results$significant_taxa
results$significant_models
which(results$p <0.05)
results$p[19]
results$all_models[19]

#wtih covariates
results <- differentialTest(formula = ~ stage_cat + Sex,
                            phi.formula = ~ stage_cat + Sex,
                            formula_null = ~ 1,
                            phi.formula_null = ~ stage_cat + Sex,
                            data = phylo,
                            test = "Wald", boot = FALSE,
                            fdr_cutoff = 0.1)

results$p_fdr
results$significant_taxa
results$significant_models
which(results$p <0.05)
results$p[7]
results$all_models[7]


#ALDEx2 DA analysis
#install.packages("BiocManager")
#BiocManager::install("ALDEx2")
library(ALDEx2)

#create model containing the variables to be tested.
matrixmodel <- model.matrix(~stage_cat, filtered_metadata) 
View(matrixmodel)
  #or with covariates
matrixmodel <- model.matrix(~stage_cat + Sex, filtered_metadata) 
View(matrixmodel)

#generate Monte Carlo samples of the Dirichlet distribution, performs centred log-ratio transformation.
CLR <- aldex.clr(rare_filter_table, matrixmodel, mc.samples = 128)
#calculates the expected values for each coefficient of a glm model on the data returned by aldex.clr
GLM <- aldex.glm(CLR, matrixmodel)
View(GLM)



```


##stacked barchart

#for taxa file - species as columns and your samples as rows

```{r}
library(ggplot2)
library(reshape2)
library(ggpubr)
library("dplyr")
library("ggpubr")
library("ggplot2")


#read metadata
metadata <- read_csv("metadata.csv")
rownames(metadata) <- metadata$Saliva_sampleID
View(metadata) 

#read taxa file
taxa_table <- read.csv("freature-table_top10_genera.csv", header = TRUE)

taxa_table <- read_csv("~/Desktop/LungCancer_Study_March2025/Saliva_analysis_July2025/feature-table_solobacterium.csv")
taxa_table <- as.data.frame(taxa_table)
rownames(taxa_table) <- taxa_table[,1]
taxa_table <- as.data.frame(taxa_table[,-c(1:1)])
taxa_table <- as.data.frame(taxa_table[-c(110:111),])
View(taxa_table)

taxa_table <- t(taxa_table)


#create taxa table with only samples of interest
#(filter taxa to match samples in the metadata file).
intersect(rownames(taxa_table), rownames(metadata))
samples_to_keep <- intersect(rownames(taxa_table), rownames(metadata))
filtered_taxa <- taxa_table[samples_to_keep,]
filtered_taxa <- as.data.frame(filtered_taxa[,-c(1:1)])
View(filtered_taxa)

RA <- sweep(filtered_taxa, 1,rowSums(filtered_taxa), "/")

rowSums(RA)
colnames(RA)


#convert data frame from a "wide" format to a "long" format
RA_melt = melt(RA, id = c("names"))

#pick colors
colours = c( "#A54657", "#86BBD8", "#33658A", "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E" ,"#E6AB02", "#A6761D", "#666666")


#make the plot
stacked_boxplot = ggplot(RA_melt, aes(x = names, fill = variable, y = value*100)) + 
    geom_bar(stat = "identity", colour = "black") + 
    theme(axis.text.x = element_text(angle = 90, size = 14, colour = "black", vjust = 0.5, hjust = 1, face= "bold"), 
    axis.title.y = element_text(size = 16, face = "bold"), legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 12, face = "bold", colour = "black"), 
    axis.text.y = element_text(colour = "black", size = 12, face = "bold")) + 
    scale_y_continuous(expand = c(0,0)) + 
    labs(x = "Sample", y = "Relative Abundance (%)", fill = "Genus") + 
    scale_fill_manual(values = colours)
stacked_boxplot


#Plot RA of specific taxa by cancer stage
RA <- sweep(rare_filter_table, 2, colSums(rare_filter_table), "/")
colSums(RA)

flipRA <-data.frame(t(RA))
colnames(flipRA)
identical(rownames(flipRA), rownames(filtered_metadata))

#Box plot - RA of specific taxa against by cancer variables
library(ggplot2)
library(ggpubr) 

#Example with Haemophilus.
DF <- data.frame(flipRA[ ,19],filtered_metadata$stage_cat)
RA_plot <- ggplot(data=subset(DF, !is.na(filtered_metadata.stage_cat)), aes(x=filtered_metadata.stage_cat, y=flipRA...19., fill = filtered_metadata.stage_cat)) +geom_point() + 
  stat_boxplot(geom ='errorbar', width = 0.5) +
  geom_boxplot(alpha=1, color=1, outlier.shape = NA) +
  labs(x = "Cancer Stage", y = "Relative Abundance") +
  labs(fill = "Cancer Stage") +
  scale_fill_manual(values=c("#1B9E77", "#D95F02", "#7570B3")) +
  theme_bw() +
  ylim(0,0.3)
  ggtitle("Haemophilus") +
  theme(plot.title = element_text(size = rel(1.5), face = "bold")) +
  geom_jitter(shape=16, position=position_jitter(0.2), size = 2) +
  theme(axis.title.x = element_text(size =16, face = "bold"))+
  theme(axis.title.y = element_text(size = 16, face = "bold", angle = 90))+
  theme(axis.text = element_text(size =14)) +
  theme(legend.title = element_text(size = 14, face = "bold"))+
  theme(legend.text = element_text(size = 14))
ggsave("RA_stage_Haemophilus2.pdf",  width = 6, height = 4)



```

##survival curves
#time: overall OR event free survival time in months
#status at time: censoring status 0=censored, 1=dead/event(progression/recurrence)

#Kaplan-Meier

```{r}
# install.packages(c("lubridate", "ggsurvfit", "gtsummary", "tidycmprsk"))
library("survival")
library("survminer")

#create survival objects and curves
#sruvit() function to general curve for entrie cohort
s1 <- survfit(Surv(OS_continuous, OS_censored) ~ 1, data = metadata)
str(s1)

#kaplan-meier plots with ggsurvit:survit2 functions
library(ggsurvfit)
survfit2(Surv(OS_continuous, OS_censored) ~ 1, data = metadata) |> 
  ggsurvfit(linewidth = 1.5, color = "red") +
  labs(
    x = "Months",
    y = "Overall survival probability"
    ) + 
  add_confidence_interval() +
  add_risktable(size = 6,theme = list(
        theme_risktable_default(axis.text.y.size = 16, 
                                plot.title.size = 16))) +
  ggeasy::easy_y_axis_labels_size(size = 18) +
  ggeasy::easy_x_axis_labels_size(size = 18) +
  ggeasy::easy_y_axis_title_size(size = 20) +
  ggeasy::easy_x_axis_title_size(size = 18)


#similar to above but split by sex
library(ggsurvfit)
survfit2(Surv(OS_continuous, OS_censored) ~ Sex, data = metadata) |> 
  ggsurvfit(linewidth = 1.5) +
  labs(
    x = "Months",
    y = "Overall survival probability"
    ) + 
  add_confidence_interval() +
  add_risktable(size = 6,theme = list(
        theme_risktable_default(axis.text.y.size = 16, 
                                plot.title.size = 16))) +
  ggeasy::easy_y_axis_labels_size(size = 18) +
  ggeasy::easy_x_axis_labels_size(size = 18) +
  ggeasy::easy_y_axis_title_size(size = 20) +
  ggeasy::easy_x_axis_title_size(size = 18) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  scale_fill_manual(values = c("#0072B2", "#D55E00"))


##Survival Analysis - Cox Hazard Ratio
#Run loop for on all taxa.
#unadjusted model
for (i in 134:241){
    cox_sum <-summary(coxph(Surv(OS_continuous, OS_censored) ~ metadata[,i], data = metadata))
  print(cat(colnames(metadata[i]),",", cox_sum$waldtest[3],",", cox_sum$conf.int[1],",",cox_sum$conf.int[3],",",cox_sum$conf.int[4] ))
  }

#adjusted model
for (i in 21:24){
    cox_sum <-summary(coxph(Surv(OS_continuous, OS_censored) ~ metadata[,i]+
                   age_at_surgery + Sex + Smoking_Status + stage_cat +
                   abx_6mon_prior + adjuvant, data = metadata))
  print(cat(colnames(metadata[i]),",", cox_sum$waldtest[3],",", cox_sum$conf.int[1],",",cox_sum$conf.int[3],",",cox_sum$conf.int[4] ))
}


# Generat Forest plots to show Cox HR results
#install.packages("forestplot")
library(forestplot)

base_data <- tibble::tibble(mean  = c(2.30, 1.20,1798.10,6.79,5.55,1.39,1.92, 4.32, 2.14,35.66,69.26),
                            lower = c(1.22, 1.02, 5.05,1.57,1.23,1.02,1.05,1.12,1.02,1.24,1.41),
                            upper = c(4.34,1.42,640782.60,29.44,25.02,1.89,3.54,16.63,4.47,1022.59,3408.70),
                            study = c("Megasphaera", "Bacteroidetes G3", "Tannerella", "Lancefieldella", "Dialister", "Peptostreptococcaceae G1","Bacilli G1", "Mycoplasma", "Treponema", "Peptostreptococcaceae G4", "Clostridiales F1 G1"),
                            HR = c("2.30", "1.20","1.7X10^2","6.79","5.55","1.39","1.92","4.32","2.14","35.7","69.3"),
                            pvalue = c("0.01","0.03","0.01","0.01","0.03","0.04","0.03","0.03","0.04","0.04","0.03"))
              
base_data |>
  forestplot(labeltext = c(study, HR, pvalue),
             clip = c(0.2, 6.6),
            txt_gp = fpTxtGp(ticks=gpar(cex=1.5), label = gpar(cex=1.5)),
             lwd.ci=4,
             graph.pos=2,
             xlog = TRUE) |>
  fp_set_style(box = "royalblue",
               line = "darkblue",
               summary = "royalblue") |>
  fp_add_header(study = c("", "Feature"),
                HR = c("", "HR"),
                pvalue = c("", "P-value")) |>
  fp_set_zebra_style("#EFEFEF")


#to visualize survival outcomes in relation to diversity a specific taxa, participants were be divided into present and absent groups.

#set groups in conditional dataframe
conditional <- metadata %>%
  mutate(
    taxa47_PA = case_when(
      taxa47 > 0 ~ "High",
      TRUE ~ "Low"))

#plot
fit <- survfit(Surv(OS_continuous, OS_censored) ~ taxa47_PA, data=conditional)
ggsurvplot(fit,
          pval = FALSE, conf.int = TRUE,
          font.x = 16,        # X-axis label
          font.y = 16,        # Y-axis label
          font.tickslab = 14, # X and Y axis tick labels
          font.legend = 12,   # Legend text
          risk.table = "abs_pct", # Add risk table
          risk.table.fontsize=5,
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          xlim=c(0,25),
          ylim=c(0.5,1),
          ylab="Overall Survival Probability",
          legend.title="Peptostreptococcaceae G1",
          legend.labs = c("Present", "Absent"),    # change legend labels.
          palette = c("#0072B2", "#D55E00"))
